\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{microtype}
\usepackage[hidelinks]{hyperref}
\usepackage{xcolor}

% Pending-number marker: paperA/paperB kernel results replace these on landing.
\newcommand{\TBD}[1]{\textcolor{red}{\textbf{[TBD: #1]}}}

\title{COLDPRESS: Container-Aware Encoding of Standard GGUF Quantizations}
% AUTHOR PLACEHOLDER — set to your preferred byline before submission.
\author{asystemoffields}
\date{July 2026}

\begin{document}
\maketitle

\begin{abstract}
The GGUF k-quant formats used by llama.cpp dominate local LLM inference, and their
encoders have been treated as fixed: given a weight file and calibration statistics,
\texttt{llama-quantize} chooses the stored bits, and downstream consumers accept the
result. We show that the container itself leaves large degrees of freedom unexercised,
and that a smarter encoder --- emitting files with \emph{byte-identical layout, type
map, and size} --- yields large quality gains with zero inference-time cost. COLDPRESS
combines three container-aware stages: (i)~\textbf{PERM}, function-preserving channel
permutations that choose \emph{which weights share k-quant sub-block scales},
exploiting the format's implicit assumption that adjacent channels have similar
magnitude; (ii)~\textbf{EF}, GPTQ-style Hessian error feedback encoded directly into
the committed two-level k-quant grid, with a processing order decoupled from storage
order; and (iii)~distillation of the container's continuous parameters (F32 norm
gains and fp16 superblock scales) against the full-precision teacher. On Qwen3-0.6B
at the Q2\_K operating point, the composed encoder reduces WikiText-2 perplexity by
$24.2\%$ relative to stock \texttt{llama-quantize} output at identical bytes, and by
$22.6\%$ relative to the best stock artifact of equal-or-smaller size, with consistent
gains ($18.6\%$) on a disjoint held-out corpus; a pre-registered $5\%$ bar is cleared
by every stage combination, and by the permutation stage alone --- a pure reordering
of the same bytes through the stock encoder. All artifacts load unmodified in stock
llama.cpp.
\end{abstract}

\section{Introduction}

Post-training quantization research typically proposes new numerical formats, then
asks the ecosystem to adopt new inference kernels. The most widely deployed format
family --- the GGUF k-quants consumed by llama.cpp and its many derivatives --- has
meanwhile changed slowly, and its \emph{encoder} has changed more slowly still: a
weighted rounding scheme with a per-block scale search, applied in whatever channel
order the architecture happens to present.

This paper takes the opposite approach: hold the format \emph{byte-fixed} and ask how
much quality the encoding decision itself leaves on the table. The constraint is
strict --- our artifacts have the same per-tensor type map, the same tensor byte
counts, and load in stock llama.cpp --- so every reported gain is attributable to the
choice of bits alone, and every existing deployment can consume the result with no
software change.

We find the unexercised freedom is substantial, and lives in three places:

\textbf{Storage order.} A k-quant container shares quantization scales across
consecutive runs of 16 or 32 channels, nested in 256-channel superblocks whose fp16
scale quantizes the sub-scales to 4--6 bits. Which channels share a block is an
accident of channel order --- but a transformer's function is invariant under
coordinated permutations of its internal spaces, and LLM weight matrices have
per-channel scales spanning orders of magnitude. Sorting channels so that similar
magnitudes share blocks is free at inference time, exactly function-preserving, and
directly reduces the dynamic range each low-bit code must span.

\textbf{Rounding correlation.} The stock encoder rounds each weight independently
under an activation-weighted objective. Inputs are correlated, so rounding error in
one column can be partially compensated by adjustments to columns that share
correlated inputs --- the GPTQ insight~\cite{gptq} --- but this had not been carried
into the two-level k-quant containers at their native 16-wide sub-block granularity.
We commit the container's scale hierarchy first, then choose the integer codes by an
error-feedback sweep whose processing order is decoupled from storage order; the
decoupling matters (\S\ref{sec:actorder}).

\textbf{Continuous parameters in a ``quantized'' file.} GGUF stores every
1-dimensional tensor in F32, so norm gains are free continuous parameters; and
dequantization is linear in each superblock's fp16 $(d, d_{\min})$ given fixed integer
codes, so a quantized model is exactly $W = d \otimes A - d_{\min} \otimes B$ with
millions of trainable scale parameters. Both can be distilled against the FP teacher
while the file remains byte-standard.

We also document a quantitative accident of the ecosystem: the llama.cpp calibration
(``imatrix'') collector only tracks \texttt{blk.*} matmuls, so the token embedding and
the output head --- $52\%$ of Qwen3-0.6B's parameters, including a 127\,MB Q6\_K
head --- are quantized with \emph{no activation weighting at all} in every stock
artifact. COLDPRESS supplies a proper lm-head Hessian and informed embedding column
weights within the same containers.

Our evaluation protocol was pre-registered before any measurement: matched-condition
comparisons (same model, same eval binary, same calibration text for every arm), a
primary gate of $\geq 5\%$ perplexity improvement over the \emph{best} stock artifact
at equal-or-smaller bytes, a no-Goodhart gate on a second held-out corpus, and an
operating point selected by a mechanical rule committed before the decisive baseline
numbers existed. An adversarial audit of the completed result (contamination checks,
timeline verification from commit history, alternative statistics) found no fatal
confound; we report its two substantive caveats inline (\S\ref{sec:audit}).

\section{Background: the k-quant container}
\label{sec:background}

A Q2\_K tensor stores, per 256-channel superblock along the input dimension: an fp16
scale $d$ and minimum $d_{\min}$; sixteen 4-bit sub-scales and sub-minimums (one per
16-channel sub-block); and 2-bit integer codes. Reconstruction of weight $w_{r,c}$ in
sub-block $b$ is $\hat w = d \cdot s_b \cdot q_{r,c} - d_{\min} \cdot m_b$. Q3\_K
through Q6\_K vary the widths and drop the minimum, but share the two-level structure.
Quantization presets assign types per tensor (e.g.\ the Q2\_K preset applies Q3\_K to
attention-V, FFN-down and attention-output tensors, and Q6\_K to the output head);
we always replicate the stock preset map exactly.

Two properties matter for what follows. First, \emph{block membership is storage
order}: the 16 channels sharing $s_b$ are consecutive, so a single outlier channel
forces a scale that crushes its fifteen neighbors. Second, \emph{the hierarchy is
committed independently of the codes}: given $(d, d_{\min}, s_b, m_b)$, each column's
integer codes may be chosen by any procedure whatsoever --- including one that visits
columns in an arbitrary order with cross-column error feedback.

\section{Method}

\subsection{PERM: choosing block membership}
\label{sec:perm}

For a permutation $P$ of an internal space, applied consistently to every tensor
touching that space, the network function is unchanged. For Qwen3-class models we
permute: the \emph{residual stream} (embedding columns, explicit lm-head columns,
q/k/v/gate/up input columns, o/down output rows, and all residual-space RMSNorm
gains --- tying-safe because logits $= (Ph)^\top (PE) = h^\top E$); each layer's
\emph{FFN hidden space} (gate/up output rows, down input columns; the SwiGLU
elementwise product permutes consistently); and each (layer, kv-head)'s
\emph{V$\to$O head-dim space} (value output rows and the corresponding attention-output
input columns of its query heads --- position-free because attention weights are
computed from q/k only). RoPE forbids q/k head-dim permutation. We verified exact
function preservation at $2.7\times10^{-6}$ relative logit drift (fp32 accumulation
noise) for random permutations.

Permutations are selected against the \emph{true container objective}: the
imatrix-weighted squared reconstruction error of a re-implementation of llama.cpp's
own weighted encoder, validated byte-faithful against \texttt{llama-quantize} output
($0.028\%$ median byte difference across 198 tensors), evaluated at each tensor's
exact target type from the preset map. A per-channel magnitude sort captured the
bulk of the available gain ($-9.7\%$ weighted error in the residual space, $-30.1\%$
across FFN spaces, up to $-54\%$ in late layers); local search was not required.
Calibration statistics move with the channels exactly: the imatrix file is permuted
entry-wise, so the stock encoder runs unmodified on the permuted model.

\subsection{EF: error feedback into the committed grid}
\label{sec:ef}

We collect input Hessians $H = X^\top X$ per matmul family over 131{,}072 calibration
tokens (one hooked forward pass). Per tensor: a first pass commits the container
hierarchy $(d, d_{\min}, s_b, m_b)$ exactly as the stock weighted encoder would; a
second pass chooses the integer codes column-by-column, propagating each column's
rounding error into not-yet-processed columns via the Cholesky factor of the dampened
inverse Hessian~\cite{gptq}; optionally the hierarchy is recommitted on the
error-compensated weights and the sweep repeated (we use two iterations). Because the
grid is fixed before the sweep, \emph{processing order is a free parameter}; we
process columns by descending Hessian diagonal (act-order). The embedding (a lookup,
for which cross-column feedback is not meaningful) instead receives weighted rounding
with first-layer input statistics as column weights; the output head receives full EF
under the lm-head Hessian. Both are unweighted in all stock artifacts
(\S\ref{sec:intro-imatrix-gap}).

\subsection{Distilling the container's continuous parameters}
\label{sec:distill}

Post-encoding, we distill against cached teacher top-256 log-probabilities (plus a
tail bucket) on calibration text, training only: (a)~the $\sim$65K F32 norm gains
(\textsc{norm}); or (b)~those plus every superblock's fp16 $(d, d_{\min})$ ---
$\sim$4.7M parameters --- using the exact linear form $W = d \otimes A - d_{\min}
\otimes B$ with integer products $A, B$ frozen (\textsc{e3b}). Trained scales are
written back through fp16 rounding; the file remains byte-standard. This is the GGUF
instantiation of norm tweaking~\cite{normtweaking} extended to the scale hierarchy,
and is closest in spirit to the fine-tuning stages of QuIP\#~\cite{quipsharp} and
AutoRound~\cite{autoround}, executed entirely inside the shipped container.

\section{Experimental setup}
\label{sec:setup}

\textbf{Conditions.} Qwen3-0.6B (revision \texttt{c1899de}); llama.cpp commit
\texttt{039e20a2} builds all artifacts and runs all evaluations; every arm (stock and
COLDPRESS) uses the same 3\,MB FineWeb-Edu calibration slice for its statistics
(stock: imatrix over 651K tokens; COLDPRESS: the same imatrix plus Hessians over a
131K-token subset --- richer statistics from \emph{identical data}). Primary
evaluation: WikiText-2 test perplexity (\texttt{-c 512}, full file), cross-domain
from calibration and touched by no pipeline stage (verified: $0.0\%$ shingle
containment against calibration, and by code path). Secondary (no-Goodhart) gate:
a FineWeb-Edu held-out slice, doc-disjoint from calibration ($0.12\%$ shingle
containment, boilerplate only). All gate comparisons are computed within a single
GPU kernel run (same VM, same binary).
% paperA refreshes: single-kernel measurement of ALL headline arms incl. IQ3_XXS.

\textbf{Pre-registration.} The claim shape, gates (G1: $\geq5\%$ WikiText ppl
improvement over the best stock artifact at $\leq$ bytes$\times1.001$; G2: $\geq2.5\%$
on the held-out corpus vs the same baseline), and kill criteria were locked before any
measurement. The operating point was selected by a mechanical rule (committed to
version control while baseline evaluation was still in progress; see \S\ref{sec:audit}
for the audit's precise timeline finding) applied to stock arms only: it selected the
Q2\_K point (347.3\,MB), whose best equal-or-smaller stock competitor is IQ3\_XXS
(345.9\,MB).

\section{Results}
\label{sec:results}

\begin{table}[t]
\centering
\small
\begin{tabular}{lrrr}
\toprule
Arm (Q2\_K map unless noted) & MB & WikiText-2 & FineWeb-held \\
\midrule
f16 reference & 1509.3 & 21.47 & 21.12 \\
\midrule
stock Q2\_K (imatrix) & 347.3 & 44.41 & 38.90 \\
stock IQ3\_XXS (imatrix; best $\leq$ bytes) & 345.9 & 43.47 & 36.99 \\
\midrule
PERM only (stock encoder) & 347.3 & 40.57 & 38.22 \\
EF only (act-order) & 347.3 & 33.97 & 30.47 \\
PERM + EF & 347.3 & \textbf{33.67} & \textbf{30.12} \\
PERM + EF + \textsc{norm} & 347.3 & \TBD{paperA} & \TBD{paperA} \\
PERM + EF + \textsc{e3b} & 347.3 & \TBD{paperA} & \TBD{paperA} \\
\bottomrule
\end{tabular}
\caption{Perplexity at the pre-registered operating point. All COLDPRESS arms have
byte-identical type maps to stock Q2\_K (asserted per tensor in-kernel). PERM+EF:
$-24.2\%$ vs stock Q2\_K and $-22.6\%$ vs IQ3\_XXS on WikiText-2; $-18.6\%$ vs
IQ3\_XXS on the held-out corpus.}
\label{tab:main}
\end{table}

Table~\ref{tab:main} gives the main result. Paired bootstrap over evaluation chunks
(10k resamples): EF-only vs IQ3\_XXS is $-15.19\%$ $[-16.11, -14.26]$ on WikiText-2
and $-10.78\%$ $[-11.70, -9.85]$ on the held-out corpus --- both pre-registered gates
are cleared at $3\times$ and $4\times$ their bars respectively, before the permutation
and distillation stages are added. A per-chunk sign test gives $518/583$ chunks
improved ($p \approx 6\times10^{-89}$); moving-block bootstraps at all tested block
sizes leave the headline conclusions unchanged.
% paperA refreshes the composed-arm CIs within one kernel.

\textbf{PERM alone} improves WikiText-2 by $8.64\%$ $[-9.47, -7.82]$ over stock Q2\_K
at identical bytes through the \emph{stock encoder} --- the artifact is produced by
unmodified \texttt{llama-quantize} on a permuted-but-equivalent model. Against
IQ3\_XXS the point estimate is $-6.67\%$; an i.i.d.\ bootstrap places the interval at
$[-7.66, -5.68]$, but a conservative 50-chunk moving-block bootstrap widens the upper
edge to $-4.8\%$, so we report PERM-alone as clearing the $5\%$ bar in point estimate
and marginally at the interval level. Its held-out improvement over stock Q2\_K is
$1.8\%$; PERM's strength is greatest exactly where containers are most starved.

\textbf{Generation quality.} At these bit rates the stock artifact is qualitatively
degraded: on a fixed-seed spot prompt, stock Q2\_K degenerates into a repetition loop
while the COLDPRESS artifact answers the question. Extended task-level evaluation
(HellaSwag, Winogrande, ARC, MMLU, and KL-divergence to the f16 teacher) is reported
in \S\ref{sec:battery}.
% evalX kernel populates \S\ref{sec:battery}.

\subsection{The composition failure, and why order matters}
\label{sec:actorder}

Our first PERM+EF composition was catastrophic: WikiText-2 perplexity 67.1, worse
than stock, while every per-tensor error metric looked unremarkable. The mechanism is
an interaction invisible to diagonal metrics: GPTQ-style sweeps shovel accumulated
error into whichever columns are processed \emph{last}, and the magnitude-sort
permutation had just arranged storage order so that the most important channels come
last. (EF simultaneously \emph{raises} diagonal-weighted error --- median $+52\%$
per tensor --- while cutting real perplexity $17\%$: it deliberately trades diagonal
error for correlated compensation, so diagnostics that look only at diagonal error
are doubly misleading here.) Decoupling processing order from storage order fixes the
composition and improves EF itself: act-order EF gains a further $7.8\%$ over
storage-order EF even in unpermuted models, and the composed PERM+EF becomes the best
arm. We believe this interaction has not been previously reported, because these two
techniques had not previously been composed.

\subsection{Mechanism attribution and point generality}
\label{sec:attribution}

Because stock artifacts quantize the embedding and output head unweighted
(\S\ref{sec:intro-imatrix-gap}), part of EF's gain may be attributable to fixing
those two tensors rather than to cross-column error feedback per se. Ablations
isolating the two (EF on \texttt{blk.*} only; upgraded embedding/head only) and EF at
the unchosen Q3\_K\_S operating point: \TBD{paperB}.

\subsection{Negative results}
\label{sec:negative}

(1)~The IQ codebook family collapses on this small model: IQ3\_XXS at 3.06 bits/weight
barely beats Q2\_K at 2.63, and IQ2\_M (73.7) is far worse than Q2\_K --- k-quants
dominate the small-model low-bit regime, which is precisely the regime our containers
target. (2)~Permutations optimized against the Q2\_K container map are approximately
neutral when applied at Q3\_K\_S ($-0.5\%$ WikiText, $+3.8\%$ held-out) --- the
grouping win is container-specific, as the mechanism predicts. (3)~Diagonal-weighted
error cannot rank EF encoders (\S\ref{sec:actorder}).

\subsection{Task battery}
\label{sec:battery}
\TBD{evalX: HellaSwag / Winogrande / ARC-e / ARC-c / MMLU accuracies and
KL-divergence-to-f16 (mean KLD, RMS $\Delta p$, same-top-1 rate) for all arms.}

\subsection{Audit}
\label{sec:audit}

An adversarial audit of the completed result verified: no calibration--evaluation
contamination (computational shingle checks); identical calibration inputs across
arms; byte-parity and type-map identity per tensor; exact reproduction of all reported
CIs from logs; robustness of the headline under sign tests and block bootstraps; and
that the champion encoder predates any COLDPRESS perplexity measurement (no
tune-on-eval loop, verified from commit history). Two caveats it required us to state:
the operating-point rule, while mechanical over stock arms only, was committed while
baseline evaluation was in progress (a 96-of-583-chunk partial of the decisive
baseline was in the repository at commit time; the completed numbers moved the
decision inputs by $<1.5\%$ and the gates were ultimately cleared at $2$--$4\times$
their bars); and the PERM-alone interval caveat reported in \S\ref{sec:results}.
% paperA replaces the cross-kernel caveat: all headline arms re-measured in one kernel.

\section{Cost and scaling}

The recipe is one-time per model: one calibration forward pass (Hessians + teacher
cache), a permutation search that is linear in parameters (a sort sufficed), an EF
sweep of the same complexity class as any GPTQ run, and a short distillation of
$\sim$5M scale parameters --- roughly one GPU-hour at 0.6B, an estimated 2--6 GPU-hours
at 7--9B. The artifact costs nothing extra at inference, forever. Known scaling
considerations: absolute low-bit headroom shrinks as models grow (margins will
compress); the residual-space permutation is shared across all layers (per-layer FFN
and V$\to$O spaces, where most of the objective gain concentrated, scale
independently); hidden dimensions must be 256-divisible (nearly all modern models);
and untied larger models additionally admit rotation-based reparameterizations that
tied small models forbid. Scaling experiments on Qwen3.5-9B and Gemma-4-12B ---
including derived permutation-space maps handling their architectural novelties
(hybrid linear-attention layers with attention output gates; dual attention
geometries with K$=$V weight sharing) --- are in progress. \TBD{scale results}

\section{Related work}

GPTQ~\cite{gptq} introduced Hessian error feedback; IST-DASLab's gptq-gguf-toolkit
and Intel AutoRound~\cite{autoround} target GGUF export, the latter with
container-native scale tuning. Our EF differs in the committed-grid formulation at
native 16-wide sub-block granularity, the act-order/storage-order decoupling, and
coverage of the tensors stock pipelines leave unweighted. Permutation-based
quantization grouping appeared in 2026 for other regimes --- PermuQuant~\cite{permuquant}
(diffusion models, NVFP4 scale groups) and CHAMP-Q~\cite{champq} (W4A8, 128-wide
groups) --- with older roots in activation-side reordering (RPTQ~\cite{rptq}) and
GPTQ's act-order. To our knowledge the k-quant sub-block instantiation, the
tied-embedding whole-model treatment, the exact imatrix permutation, and the
weight-side permutation of a shipped standard container are new. Rotation methods
(QuaRot~\cite{quarot}, SpinQuant~\cite{spinquant}) reparameterize more aggressively
but break embedding tying on small models ($+50\%$ bytes here) and were rejected at
design time; norm tweaking~\cite{normtweaking} is the ancestor of our \textsc{norm}
stage; QuIP\#~\cite{quipsharp} and AQLM~\cite{aqlm} achieve strong 2-bit results in
non-standard containers requiring custom kernels.

\section{Limitations}

Results are established at one scale (0.6B) and one operating point chosen by our
pre-registered rule; scale replication is in progress. The held-out gate corpus
shares a domain with calibration (WikiText-2 provides the cross-domain check).
PERM-alone clears the primary bar marginally under conservative statistics. The
strongest single stage (EF) is the least novel; the novel stages (PERM, scale
distillation) contribute measurably but the composition carries the headline.
Gains concentrate at $\leq$3 bits/weight where stock quality is most degraded ---
translated conservatively, the composed artifact performs like a stock artifact
roughly $5$--$6\%$ larger.

\section{Reproducibility}

Everything is deterministic from pins: llama.cpp commit, model revision, calibration
slice, permutation file, and self-contained kernel scripts that rebuild all inputs
(repository: \TBD{repo URL}). Verification gates ship with the pipeline: exact
function preservation for permutations, byte-faithfulness of the encoder
re-implementation against \texttt{llama-quantize}, per-tensor type-map identity
asserts, and round-trip identity of the GGUF writer.

\section{Conclusion}

Holding a deployed format byte-fixed and re-deciding only the bits recovers double-digit
perplexity improvements at zero inference cost: block membership, rounding correlation,
and the container's own continuous parameters are all decision variables the ecosystem
has been leaving to chance. The artifacts drop into existing llama.cpp deployments
unchanged, the recipe is a few GPU-hours per model, and the components compose ---
once the one genuinely dangerous interaction (sorted storage order versus sequential
error dumping) is handled by construction.

This work --- the research program, the prior-art analysis, the pre-registration, the
implementation of every component, the debugging, the adversarial audit of its own
result, and this paper --- was carried out entirely autonomously by Fable 5 running in
Claude Code, over roughly one day of wall-clock time and a few dollars of shared free
compute, under a single standing instruction from its human operator. The operator
supplied the goal, the compute accounts, and two mid-course directives (use cloud
compute; write the paper); every technical decision, experiment, failure, and recovery
between those messages was the model's own. We offer this less as a claim about this
paper than as a datapoint about the near-term shape of research itself.

\begin{thebibliography}{9}
\bibitem{gptq} E.~Frantar, S.~Ashkboos, T.~Hoefler, D.~Alistarh.
GPTQ: Accurate Post-Training Quantization for Generative Pre-trained Transformers.
arXiv:2210.17323, 2022.
\bibitem{rptq} Z.~Yuan et al. RPTQ: Reorder-based Post-training Quantization for
Large Language Models. arXiv:2304.01089, 2023.
\bibitem{normtweaking} L.~Li et al. Norm Tweaking: High-performance Low-bit
Quantization of Large Language Models. AAAI 2024 (arXiv:2309.02784).
\bibitem{quarot} S.~Ashkboos et al. QuaRot: Outlier-Free 4-Bit Inference in Rotated
LLMs. arXiv:2404.00456, 2024.
\bibitem{spinquant} Z.~Liu et al. SpinQuant: LLM Quantization with Learned
Rotations. arXiv:2405.16406, 2024.
\bibitem{quipsharp} A.~Tseng et al. QuIP\#: Even Better LLM Quantization with
Hadamard Incoherence and Lattice Codebooks. arXiv:2402.04396, 2024.
\bibitem{aqlm} V.~Egiazarian et al. Extreme Compression of Large Language Models via
Additive Quantization. arXiv:2401.06118, 2024.
\bibitem{autoround} W.~Cheng et al. Optimize Weight Rounding via Signed Gradient
Descent for the Quantization of LLMs. arXiv:2309.05516; AutoRound GGUF export, Intel, 2024--2026.
\bibitem{permuquant} PermuQuant: Permutation-Aware Group Quantization.
arXiv:2605.09503, 2026.
\bibitem{champq} CHAMP-Q: Channel-Aligned Magnitude Permutation for W4A8
Quantization. 2026.
\end{thebibliography}

\end{document}
